A model-based method for detecting persistent cultural change using panel data[^thanks]

### 2021-02-08 ms.Rmd
conditional accept edits

@@

>>>>

This model constrains $\rho$ to be equal across waves. This implies two important assumptions. First, it assumes that the time between waves is equal. This is a reasonable assumption for the GSS, where surveys take place about two years apart. Across the three panels, the average time between waves 1 and 2 is between 695 and 764 days, while the average time between waves 2 and 3 is between 692 and 743 days. This might not be a reasonable assumption in other panels and can be loosened by adding a coefficient to $\rho$ for the time since last wave.

Second, constraining $\rho$ to be the same across waves assumes that active updating happens by a linear secular trend. While this is likely a valid assumption for many questions we explore in the GSS, there are questions where opinions likely change discontinuously and in durable ways around specific dates and events. In the GSS panels, these questions include confidence in various branches of the government, which might change as a result of elections or high-profile political events, and confidence in financial institutions, belief in the ability to find a job, and subjective socioeconomic status, which might change in response to the onset of the Great Recession.[^valid]

In our framework, $\rho$ should be thought of as a detector of whether any active updating happens in the panel window, rather than as a measure of the rate of active updating over time. If any updating happens in the panel window, $\rho$ will tend to be non-zero. However, we cannot rule out the possibility that all attitudes can demonstrative durable updating under certain circumstances, only whether they demonstrate updating in our study window. We return to this point in the discussion below.

[^valid]: Despite the (potentially erroneous) assumption of linear updating, these questions still prefer the active updating model in relevant windows in the analysis below.

@@

As noted above, the fact that most attitudes prefer the settled dispositions model in the panel window should not be taken to imply that settled attitudes cannot durably update if circumstances change. Consistent with previous work, the results presented here suggest that circumstances occasionally change and lead to durable attitude change. At the same time, our results suggest that the kinds of circumstantial change that lead to durable attitude change seem relatively rare, and seem confined to major public events. If local changes such as changes in social network composition do lead to durable updating in attitudes, these local changes are so rare as to be undetectable on most issues measured here.
